PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-04-20 papers/NAR_web_server_2009/paper.sup.tex
Manuscript changes

To calculate this MST we use Prim's algorithm \cite{Prim57}.  For a
graph of $n$ nodes and $e$ edges, Prim's can be implemented such that
the algorithm has an $O(e + n \log n)$ worst-case runtime.  However,
the {\it narrow-band} nature of the model allows us to reduce this
bound.  If we were to model only the edges connecting adjacent nodes,
we would find the node degrees to be very small relative to $n$: any
EST overlaps only a few others, and has no connection to a vast
majority of the data set members.  Since we can quickly eliminate most
of these excess edges with the $u/v$ and $t/v$ heuristics (removing
more than 99\% of all edges before applying Prim's), we find in
practice that $e$ is a very small fraction of $n^2$.  In the
runtime results (Figure~4), we find that when holding the EST size
distribution constant, the tool as a whole has a runtime of $O(n^2)$
-- appearing to dominated by the time required to apply
the filtering heuristics to every EST pair.

>>>>

To calculate this MST we use Prim's algorithm \cite{Prim57}.  For a
graph of $n$ nodes and $e$ edges, Prim's can be implemented such that
the algorithm has an $O(e + n \log n)$ worst-case runtime.  However,
the {\it narrow-band} nature of the model allows us to reduce this
bound.  If we were to model only the edges connecting adjacent nodes,
we would find the node degrees to be very small relative to $n$: any
EST overlaps only a few others, and has no connection to a vast
majority of the data set members.  Since we can quickly eliminate most
of these excess edges with the $u/v$ and $t/v$ heuristics (removing
more than 99\% of all edges before applying Prim's), we find in
practice that $e$ is a very small fraction of $n^2$.  In the
runtime results (Figure~4), we find that when holding the EST size
distribution constant, the tool as a whole has a runtime of $O(n^2)$
-- appearing to be dominated by the time required to apply
the filtering heuristics to every EST pair.

### 2010-04-20 papers/NAR_web_server_2009/paper.tex
Manuscript changes

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a clustering in the process of assembly).  For
our simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03}
to generate simulated transcript fragments of varying length under
different models of error (Supplementary Materials, Section~D.1),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
to match, as closely as possible, those used in the \wcd\/ study (see
Supplementary Materials).  \mc{The most important method of quality
  assessment is {\it sensitivity} (the fraction of fragment pairs from
  the same gene that were} \mc{correctly clustered together).  But we also
  look at the {\it Jaccard Index} (which balances sensitivity with the
  number of false positives by looking at the ratio of true positives
  to the sum of true positives, false positives, and true negatives),
  {\it Type 1 error} (the fraction of genes that were divided between
  clusters), and {\it Type 2 error} (the fraction of clusters
  containing two or more genes).  Specificity was not an issue in
  simulated tests, as both \peace\/ and \wcd\/ were completely
  successful in separating unrelated ESTs (save when dealing with
  recently duplicated genes).  In Figure~\ref{SeJiT1T2} we plot
  these four tests as a function of error rate, and observe the almost
  identical results between \peace\/ and \wcd\/.  In
  Figure~\ref{seq_parallel} we plot the runtime for \peace\/ and
  \wcd\/, again observing almost identical results when run
  sequentially -- but significantly faster runtime for PEACE on
  multiple processor when holding the EST/processor ratio constant
  (ranging from a 65\% improvement for two processors too a 17\%
  improvement for 12 processors).}

@@

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error, based on the average over 30 simulated Sanger
    Sequence ESTs sets derived from 100 zebra fish genes (see
    Supplementary Materials, Section C, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd \mc{(version 0.5.1)}, Black/Dot-Dash =
    \capthree; vertical tics = 95\% confidence intervals on estimates.
    Intervals are not presented for Type 1 error due to the effective
    lack of variance.}\label{SeJiT1T2}
\end{figure}

\begin{figure}
   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.pdf}}
   \caption{Comparisons of Runtime: On left, we compare the sequential
     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
     runtime is not reported, as the time spent on clustering cannot
     be differentiated from the time spent on assembly.)  On right,
     a comparison when run in parallel, ranging the number of
     processors from 2 to 12 while holding the EST/Processor ratio
     steady at the constant 8888.  All values represent the average
     of 30 runs; vertical tics = 95\% confidence intervals on
     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
     cache and 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
\end{figure}

\mc{In applying the tools to real data Sanger data, we used the Human
Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
\cite{Hazelhurst08a,Hazelhurst08b}, (see Table~\ref{SangerReal}).  We notice essentially identical
results for PEACE and WCD in quality, both significantly better than
Cap3 in Sensitivity and Type 1 error rate, while slightly worse in the
Jaccard Index and Type 2 error.  In runtime we see some inconsistency,
with PEACE showing a 60\% runtime improvement in the first data set,
but requiring 20\% more time in the Arabidopsis data set.  In Section~D.1
of the Supplementary Materials we present runtimes several more
sets, observing that while \peace\/ appears to be significantly more
faster on the smaller sets, \wcd\/ does overtake it for larger sets.}

\begin{table*}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline Bicyclist Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\hline
\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
  standard quality measurements.}\label{SangerReal}
\end{table*}

@@

@@

\mc{Here we have presented \peace, a stand alone tool for the
high-throughput clustering of transcript fragments capable of dealing
with sequences as short as 50 bases.  \peace\/, available at
\href{http://www.peace-tools.org}{www.peace-tools.org}, is open-source
and managed through a user friendly GUI that enables both local and
remote installation and execution in sequential or parallel mode.
Based on a novel algorithm for the clustering of the fragments by gene
association, \peace\/ shows significant improvement in sensitivity,
over the competing \wcd\/ tool \cite{Hazelhurst08a} when applied to
NGS reads, matches \wcd\/ when applied to Traditional Sanger
Sequencing output, and shows an order of magnitude in improvement over
the clustering performed in the course of assembly by the \capthree\/
tool \cite{Huang99}.}

@@

>>>>

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a clustering in the process of assembly).  For
our simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03}
to generate simulated transcript fragments of varying length under
different models of error (Supplementary Materials, Section~D.1),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
to match, as closely as possible, those used in the \wcd\/ study (see
Supplementary Materials).  \mc{The most important method of quality
  assessment is {\it sensitivity} (the fraction of fragment pairs from
  the same gene that were} \mc{correctly clustered together).  But we
  also look at the {\it Jaccard Index} (which balances sensitivity
  with the number of false positives by looking at the ratio of true
  positives to the sum of true positives, false positives, and true
  negatives), {\it Type 1 error} (the fraction of genes that were
  divided between clusters), and {\it Type 2 error} (the fraction of
  clusters containing two or more genes).  Specificity was not an
  issue in simulated tests, as both \peace\/ and \wcd\/ were
  completely successful in separating unrelated ESTs (with the
  exception of recently duplicated genes, as discussed in the
  Supplementary Materials).  In Figure~\ref{SeJiT1T2} we plot these
  four tests as a function of error rate, and observe the almost
  identical results between \peace\/ and \wcd\/.  In
  Figure~\ref{seq_parallel} we plot the runtime for \peace\/ and
  \wcd\/, again observing almost identical results when run
  sequentially -- but significantly faster runtime for PEACE on
  multiple processor when holding the EST/processor ratio constant
  (ranging from a 65\% improvement for two processors too a 17\%
  improvement for 12 processors).}

@@

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.png}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error, based on the average over 30 simulated Sanger
    Sequence ESTs sets derived from 100 zebra fish genes (see
    Supplementary Materials, Section C, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd\/ \mc{(version 0.5.1)}, Black/Dot-Dash =
    \capthree; vertical tics = 95\% confidence intervals on estimates.
    Intervals are not presented for Type 1 error due to the effective
    lack of variance.}\label{SeJiT1T2}
\end{figure}

\begin{figure}
   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.png}}
   \caption{Comparisons of Runtime: On left, we compare the sequential
     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
     runtime is not reported, as the time spent on clustering cannot
     be differentiated from the time spent on assembly.)  On right,
     a comparison when run in parallel, ranging the number of
     processors from 2 to 12 while holding the EST/Processor ratio
     steady at the constant 8888.  All values represent the average
     of 30 runs; vertical tics = 95\% confidence intervals on
     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
     cache and 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
\end{figure}

\mc{In applying the tools to real data Sanger data, we used the Human
  Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
  the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
  \cite{Hazelhurst08a,Hazelhurst08b}, (see Table~\ref{SangerReal}).
  We notice essentially identical results for PEACE and WCD in
  quality, both significantly better than Cap3 in Sensitivity and Type
  1 error rate, while slightly worse in the Jaccard Index and Type 2
  error.  In runtime we see some inconsistency, with PEACE showing a
  60\% runtime improvement in the first data set, but requiring 20\%
  more time in the Arabidopsis data set.  In Section~D.1 of the
  Supplementary Materials we present runtimes several more sets,
  observing that while \peace\/ appears to be significantly 
  faster on the smaller sets, \wcd\/ does overtake it for larger
  sets.}

\begin{table*}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline EasyCluster Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\hline
\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
  standard quality measurements.}\label{SangerReal}
\end{table*}

@@

\begin{table*}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline EasyCluster Human & \peace & 0.871 & 0.809 &
\cline{2-9} Benchmark & \wcd &  0.020 & 0.020 &
\cline{2-9} (111 Genes) & \capthree & 0.005 & 0.004 &
\hline
\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
  standard quality measurements.}\label{SangerReal}
\end{table*}

@@

\mc{Here we have presented \peace, a stand alone tool for the
  high-throughput clustering of transcript fragments capable of
  dealing with sequences as short as 50 bases.  \peace\/, available at
  \href{http://www.peace-tools.org}{www.peace-tools.org}, is
  open-source and managed through a user friendly GUI that enables
  both local and remote installation and execution in sequential or
  parallel mode.  Based on a novel minimum-spanning-tree based
  algorithm for the clustering of the fragments by gene association,
  \peace\/ shows significant improvement in sensitivity, over the
  competing \wcd\/ tool \cite{Hazelhurst08a} when applied to NGS
  reads, matches \wcd\/ when applied to Traditional Sanger Sequencing
  output, and shows an order of magnitude in improvement over the
  clustering performed in the course of assembly by the \capthree\/
  tool \cite{Huang99}.}

@@

Peace can be downloaded from
\href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
committed to maintaining and improving the tool.  The underlying PEACE
architecture is configured to support a variety of tasks, including the
possibility of a minimum spanning tree based assembly tool we hope to
add to the tool in a future study.
